The First Level of Super Mario Bros.~is Easy with Lexicographic Orderings and Time Travel {\normalsize \ldots after that it gets a little tricky.}

### 2013-03-17 tasbot/paper/paper.tex
Caching, compression, ...

\newcommand\comment[1]{}
\newcommand\parameteralert[1]{
  % XXX this can get cooler.
  \begin{quotation}
  {\bf Parameter Alert!}\,\, #1
  \end{quotation}
}

@@

\begin{figure}[h!tb]
\begin{center}
\includegraphics[width=0.95 \linewidth]{mario12}
\end{center}\vspace{-0.1in}
\caption{This is where it starts to get hard, even with lexicographic
ordering and time travel. This overhang is very tight in terms of the
inputs that can get you through it; random play tends to kill you because
of the advancing enemies that can't easily be jumped over. Greedy
algorithms can't resist the ledge with the coins, but then need to
turn around.}
\label{fig:mario12}
\end{figure}

@@

If they all do, then the score for the improvement is 
$\int_s^n - \int_s^o + (n - o)$, which is guaranteed to be positive.

@@

\paragraph{Takes of woe, tales of joy.} 
Backtracking was initially added in order to fix the mario coin ledge
problem in {\tt 1-2}. It did not help with this. Older versions of the
code would have Mario get past this spot, probably by luck, but in
these he would get stuck in an embarrassing way on a small pit later,
or jump straight into an enemy. Most of the approaches that actually
looked solidly good elsewhere were failing badly here. As I started
writing up this paper, on an airplane, I discovered the bug in the
weighting of objective functions described in
Section~\label{sec:objectiveweighting}. On my trip I had let {\tt
  playfun} run with particularly high constants ({\tt NFUTURES} = 50,
lots of backtracking candidates, etc.) and it had spent thousands of
CPU hours playing to this point, grabbing the coins, and then dying of
timeout, then losing all its lives, starting over, and getting to this
point again! After fixing the bug, I tried again, and it was a huge
breakthrough: Mario jumped up to get all the coins, and then almost
immediately jumped back down to continue the level! On his first try
he beat {\tt 1-2} and then immediately jumped in the pit at the
beginning of {\tt 1-3} just as I was starting to feel suuuuuper smart
(Figure~\ref{fig:mario13}). On a scale of OMFG to WTFLOL I was like
{\em whaaaaaaat}?

\begin{figure}[ht]
\begin{center}
\includegraphics[width=0.95 \linewidth]{adventureisland}
\end{center}\vspace{-0.1in}
\caption{After making me feel so successful by finally cleanly
  finishing world {\tt 1-2}, Mario immediately jumps into the first
  trivial pit in {\tt 1-3}. I believe what's going on here is that
  this is actually the best future, because resetting back a single
  screen isn't a large loss, and probably the rest of the futures were
  struggling to make progress. This level is very jumpy and probably
  needs more futures to navigate its obstacles. In any case, Mario,
  Touch\'e.}
\label{fig:mario13}
\end{figure}

Seeing this huge improvement changed my idea about what part of the
code needed work (I now believe that simpler search strategies might
work and better lexicographic order generation and weighting is called
for). But, this was already in the midst of writing up the paper, so
instead I spent the CPU time running the current version on a bunch of
games. Those results are described in Section~\ref{sec:results} and
videos are available on \url{http://tom7.org/mario/}. In the next
section I describe some of the performance improvements I made, and
other modifications to the emulator interface.

@@

@@

@@

@@

>>>>

\newcommand\comment[1]{}
\newcommand\parameteralert[1]{
  % XXX this can get cooler.
  \begin{quotation}
  \noindent {\bf Parameter Alert!}\, #1
  \end{quotation}
}

@@

\begin{figure}[h!tb]
\begin{center}
\includegraphics[width=0.95 \linewidth]{mario12}
\end{center}\vspace{-0.1in}
\caption{This is where it starts to get hard, even with lexicographic
ordering and time travel. This overhang is very tight in terms of the
inputs that can get you through it; random play tends to kill you because
of the advancing enemies that can't easily be jumped over. Greedy
algorithms can't resist the ledge with the coins, but then would need to
turn around.}
\label{fig:mario12}
\end{figure}

@@

The integral has to be at least as good as before, the new integral
has to be positive, and the new end state needs to look better than
the old end state (the triangle inequality does not necessarily hold
here). If they all do, then the score for the improvement is $\int_s^n
- \int_s^o + (n - o)$, which is guaranteed to be positive.

@@

\paragraph{Takes of woe, tales of joy.} 
Backtracking was initially added in order to fix the mario coin ledge
problem in {\tt 1-2}. It did not help with this. Older versions of the
code would have Mario get past this spot, probably by luck, but in
these he would get stuck in an embarrassing way on a small pit later,
or jump straight into an enemy. Most of the approaches that actually
looked solidly good elsewhere were failing badly here. As I started
writing up this paper, on an airplane, I discovered the bug in the
weighting of objective functions described in
Section~\ref{sec:objectiveweighting}. On my trip I had let {\tt
  playfun} run with particularly high constants ({\tt NFUTURES} = 50,
lots of backtracking candidates, etc.) and it had spent thousands of
CPU hours playing to this point, grabbing the coins, and then dying of
timeout, then losing all its lives, starting over, and getting to this
point again! After fixing the bug, I tried again, and it was a huge
breakthrough: Mario jumped up to get all the coins, and then almost
immediately jumped back down to continue the level! On his first try
he beat {\tt 1-2} and then immediately jumped in the pit at the
beginning of {\tt 1-3} just as I was starting to feel suuuuuper smart
(Figure~\ref{fig:mario13}). On a scale of OMFG to WTFLOL I was like
{\em whaaaaaaat}?

\begin{figure}[ht]
\begin{center}
\includegraphics[width=0.95 \linewidth]{mario13}
\end{center}\vspace{-0.1in}
\caption{After making me feel so successful by finally cleanly
  finishing world {\tt 1-2}, Mario immediately jumps into the first
  trivial pit in {\tt 1-3}. I believe what's going on here is that
  this is actually the best future, because resetting back a single
  screen isn't a large loss, lives aren't included in the objective
  function, and probably the rest of the futures were struggling to
  make progress. This level is very jumpy and probably needs more
  futures to navigate its obstacles. In any case, Mario, Touch\'e.}
\label{fig:mario13}
\end{figure}

Seeing this huge improvement changed my idea about what part of the
code needed work (I now believe that simpler search strategies might
work and better lexicographic order generation and weighting is called
for). But, this was already in the midst of writing up the paper, so
instead I spent the CPU time running the current version on a bunch of
games. Those results are described in Section~\ref{sec:results} and
videos are available at \url{http://tom7.org/mario/}. In the next
section I describe some of the performance improvements I made, and
other modifications to the emulator interface.

@@

Performance is really important, both because the quality of the
output is CPU-bound and because I am impatient. In its current state,
running {\tt playfun} is an overnight affair; it takes about an hour
of real time to produce 1000 output frames (16 seconds of gameplay) on
a high-end desktop machine. Other than a trick that I didn't end up
using, these performance ideas are not at all new, but they are
documented here for completeness. I certainly spent lots of time on 'em!

@@

The most expensive part, by far, is running a step of emulation. It
takes about 500\textmu s to emulate a single frame, though this can
vary depending on what instructions are actually executed (recall that
each frame corresponds to many many 2A03 instructions; there are
approximately 16,000 clock cycles per frame!). Hardware like the
Picture Processing Unit and sound chip are emulated as well, which may
actually be less efficient than the actual hardware (for example, the
sound hardware can implement filtering with passive physical
components like capacitors, but FCEUX contains a finite impulse
response filter running on IEEE floating point numbers). We want to
avoid executing a step more than once, if at all possible.

Caching is so canonical that some have called it (pejoratively) the
only idea in Systems research. And it is what we do here. The emulator
adds a call
%
\begin{code}
void CachingStep(uint8 input);
\end{code}
%
with exactly the same semantics as {\tt Step}. However, it first
checks to see if this exact input has been executed on this exact
start state before. If so, it just restores the cached result
state instead of executing the step.

I use the excellent CityHash function\cite{CityHash} as the hash code,
and use a least-recently-used approach to clean out the hash table
when it has more than a fixed {\em slop} amount more than the target
size. States are large because they contain both the savestate of the
before and after state. I have plenty of RAM, so I allow each process
to store up to 100,000 states with 10,000 states of slop, which takes
about 3 gigabytes per process.

Caching adds a little bit of overhead (a single step is about 13\%
slower, amortized), from saving the state, computing the hash code,
copying the state into the table, and cleaning the hash table
periodically. A cache hit is a zillion times faster than actually
executing the step, however, so as long as the hit rate is more than
13\%, it's worth doing. I use caching step in most places inside {\tt
  playfun}, except when I know that the state can't have been computed
yet. The {\tt playfun algorithm} is particularly suitable to caching:
The set of potential next input sequences {\tt nexts} usually share
several input prefixes, and we keep around futures for some time, so
we can often end up evaluating the same one under the same conditions
many times. Moreover, mutant futures usually share their first half,
making them half price. A really important property of caching is that
it's based on the state of memory but doesn't care about the actual
sequence used to get there. This means that in cases where the input
is ignored (Mario ignores the jump button most of the time while in
the air, for example, and ignores all inputs between levels) we reach
equivalent states and can use the cache if the next input is exactly
the same. The ideal approach here would be to follow how the bits of
the input are actually read by the code, and insert a more generic key
into the hash table. For example, if we see that the code never even
read the bit of the input corresponding to the up direction, then we
can reuse the step for an input that matches all bits but the up bit!
This of course requires a lot of mucking around in the internals,
which on a scale of within the scope to beyond the scope of this
article is a 9.9.

{\bf Software engineering lesson:} Initial results from caching were
disappointing and I thought the overhead was to blame. I made several
low-level tweaks to avoid copying, etc., to reduce the overhead from
36\% to 13\%. Then I discovered that there were 0 cache hits ever,
because of a bug (I was inadvertantly using pointer equality on keys
instead of value equality, so keys were never found). Check basic
things about your code's correctness before going on an optimization
soiree!

\subsection{Space optimizations}

Before I had the idea that is the actual subject of this paper, I was
playing around with other emulator search ideas that involved storing
very very large numbers of states. This necessitated minimal
representations of savestates. FCEUX uses zlib internally to compress
states, which works well; they shrink from something like 13,776
bytes\footnote{I don't completely understand the NES architecture and
  emulation strategy, but I believe some games include expansion chips
  that require more space for save states. All this work is based on
  the 2048 bytes of main NES RAM, as you already know. Perhaps clever
  video game authors from the past can protect against this paper's
  approach by storing important game facts inside expansion chips
  in the cartridge.} to an average of 2266. I made some modifications
to the save/load routines to make this as small as I could manage.
I removed the backing buffer, which is only used for drawing graphics,
movie info, header bytes that are constant or already known (savestate
size), which shrunk the compressed size to an average of 2047.42 bytes.
That meant I could fit about 32 million savestates in RAM at once, which
is kind of amazing.

Compression algoritms work well when there is lots of redundancy in
their input. Comparing memories between two frames in some game,
you'll often see only a few differences. Some of the memory contains
stuff that's essentially read-only, even. To improve the
compressibility of savestates, I introduce the idea of a {\em basis}
state. This is any non-empty sequence of bytes which is an additional
argument to the {\tt LoadState} and {\tt SaveState} functions. It's
subtracted from the savestate prior to compression (if it's longer or
shorter, it's treated as an infinite repetition of those bytes). If
the basis is the same as the savestate, then the result is all zeroes,
which compresses nicely (of course, now you need to have the basis in
order to load the state, which is the same as the savestate, so that
didn't save you much!). But memories during gameplay often don't
differ much, so I save a state from any arbitrary frame in the game
and then use that as the basis. This factors out any common parts
and makes the states much more compressible: The average size drops
to 1870.41 bytes.

Using a better compression algorithm like the Burrows--Wheeler
transform\cite{Burrows}\footnote{This is really one of the all-time
  great algorithms. If you don't know it and you like this kind of
  thing, you should check it out. It's so surprising and elegant.}
would probably help too; zlip implements LZW which is only mediocre.
However, on a scale of throwing my computer out the window to wanting
to reimplement all software in my own private SML library, I just
can't stand getting 3\rd-party libraries to compile and link into my
applications, so I didn't even try. In any case, I abandoned this
direction, which was not working well even with lean savestates.

For {\tt playfun}, RAM is not a bottleneck at all. I disable
compression completely on savestates, including the builtin zlib stuff
(which is actually rather slow) and just use raw savestates. The
removal of unnecessary headers and stuff is still there, and saves a
few bytes and CPU cycles, but there was no reason to actually do it
for this particular work. {\em But sometimes I do unnecessary things.}

@@

The {\tt playfun} algorithm involves running 40 or so different
10-sequence {\tt next} steps and then scoring them against {\tt
  NFUTURES} different futures. Each {\tt next} and each future is
completely independent and dominated by the cost of evaluating
emulator steps. The ideal thing would be to use threads and shared
memory to run these steps in parallel. As I mentioned earlier,
FCEUX is hopelessly single-threaded.

Eventually I realized I needed to search more states than I could in a
single thread, and so I created MARIONET. It's a play on words, a
double entendre of ``Mario network'' and ``Marionette'', which is
obvious. This is a mode whereby {\tt playfun} can be started in
``helper'' mode (listens on a port for work to do) or ``master'' mode
(connects to a list of ports to run work on), in order to run
many emulators in different processes.

@@

bug exploitation (Figure~\ref{fig:mario13bug1}).

\begin{figure}[h!tb]
\begin{center}
\includegraphics[width=0.75 \linewidth]{mario13bug1a} \\[0.3em]
\includegraphics[width=0.75 \linewidth]{mario13bug1b}
\end{center}\vspace{-0.1in}
\caption{Mario bounces off one Goomba up into the feet of another. Not
  only doesn't he have enough velocity to reach the platform, but he's
  about to hit that Goomba from below. Believe it or not, this ends
  well: {\tt playfun} is happy to exploit bugs in the game; in this case,
  that whenever Mario is moving downward (his jump crests just before the
  Goomba hits him) this counts as ``stomping'' on an enemy, even if that
  enemy is above him. The additional bounce from this Goomba also allows
  him to make it up to the platform, which he wouldn't have otherwise!
}
\label{fig:mario13bug1}
\end{figure}
